---
title: "1a. Introduction to Network Science (R / igraph edition)"
format:
  html:
    toc: true
    toc-depth: 2
    code-fold: false
    embed-resources: true
execute:
  warning: false
  message: false
---

# 1a. Introduction to Network Science

Welcome to the first practical of the Network Science Summer School!

## Learning objectives
By the end of this notebook you will be able to:

1. Recognise the most common file formats for network data and load them with `igraph`.
2. Build a network from an adjacency list (e.g. a TSV file), and a GraphML file.
3. Visualise networks using different layouts and visual encodings.
4. Describe a network with summary statistics (components, diameter, density, clustering, assortativity, robustness).
5. Plot degree, triangle, clustering and path-length distributions.
6. Compare an observed metric against random reference models (ER, configuration, Barabasi-Albert) and run a permutation test for attribute assortativity.

## Section index

- **Section 0**: Setup
- **Section 1**: Network data formats
- **Section 2**: Build a network from raw data
- **Section 3**: Visualizing networks
- **Section 4**: Describing networks
- **Section 5**: Distributions
- **Section 6**: Reference models and permutation test

## Section 0: Setup

```{r}
#| label: c-install
# If you are missing packages, install them by uncommenting:
# install.packages(c("igraph", "dplyr", "readr", "tidyr", "ggplot2", "network"))
```

```{r}
#| label: c-path
# Path to the data folder (relative to this .qmd file)
path_data <- "../../../Data/"
```

```{r}
#| label: c-setup
# Core libraries
library(igraph)     # Primary network library for this practical
library(dplyr)      # Data manipulation
library(readr)      # Reading CSV / TSV files
library(tidyr)      # Reshaping data
library(ggplot2)    # Plotting

# Visual style for the whole notebook
palette_ns <- c("#3d348b", "#e6af2e", "#191716", "#e0e2db")
theme_set(
  theme_minimal(base_size = 13) +
    theme(
      panel.grid.minor = element_blank(),
      panel.grid.major = element_line(color = "lightgray"),
      axis.text   = element_text(color = "#484848"),
      axis.title  = element_text(color = "#484848"),
      plot.title  = element_text(color = "#484848", hjust = 0),
      legend.position = "right"
    )
)

# Reproducibility seed (igraph layouts and random graphs use R's RNG)
set.seed(1)
```

---

## Section 1: Network data formats

A network is, fundamentally, just a list of who-relates-to-whom. The same network can be stored on disk in many different formats. In this section we look at the four formats you are most likely to encounter:

| Format | Extension | Best for |
|--------|-----------|----------|
| Edgelist | `.csv`, `.tsv`, `.txt` | Sharing, scraping, quick imports |
| Adjacency matrix | `.csv`, `.rds` | Linear-algebra views, small dense graphs |
| GraphML / GML / Pajek | `.graphml`, `.gml`, `.net` | Keeping node/edge attributes, interop with Gephi, Cytoscape, Python |
| JSON | `.json` | Web visualisations (d3.js, vis.js, cytoscape.js) |

### 1.1 Edgelist (CSV / TSV)

The most human-readable way to describe a network: each row is one edge.

```
source,target
Alice,Bob
Alice,Carol
Bob,Dave
...
```

In `igraph` we typically build a graph from an edgelist stored in a data frame using `graph_from_data_frame()`.

```{r}
#| label: c-toy-edgelist
# Small change to try: add one extra pair to `edges_df` and rerun.
edges_df <- tibble(
  source = c("Alice", "Alice", "Bob",  "Carol", "Dave", "Eve"),
  target = c("Bob",   "Carol", "Dave", "Dave",  "Eve",  "Frank")
)

# Showing the edge list as a data frame
edges_df

# Creating the network from an edgelist
g_toy <- graph_from_data_frame(edges_df, directed = FALSE)

# Printing the number of nodes (vcount) and edges (ecount)
cat("Nodes:", V(g_toy)$name, "\n")
cat(sprintf("n = %d, m = %d\n", vcount(g_toy), ecount(g_toy)))
```

### 1.2 Adjacency matrix view

The same network can be written as a square 0/1 matrix where `A[i, j] = 1` iff there is an edge between node `i` and node `j`. This view is convenient for linear algebra and spectral methods.

```{r}
#| label: c-toy-adjacency-sparse
## Sparse matrix (what igraph returns by default)
A <- as_adjacency_matrix(g_toy)
print(A)
```

```{r}
#| label: c-toy-adjacency-dense
# Dense adjacency matrix (small networks only!)
A <- as_adjacency_matrix(g_toy, sparse = FALSE)
print(A)

# Plot as a heatmap
image(
  seq_len(ncol(A)), seq_len(nrow(A)), t(A)[, nrow(A):1],
  col = c("white", "#3d348b"),
  axes = FALSE, xlab = "", ylab = "",
  main = "Adjacency matrix of the toy network"
)
axis(1, at = seq_len(ncol(A)), labels = colnames(A), las = 2)
axis(2, at = seq_len(nrow(A)), labels = rev(rownames(A)), las = 1)
```

### 1.3 Interchange formats: GraphML, GML, Pajek, JSON

When you want to keep **node and edge attributes** (labels, weights, types, positions, ...) alongside the edges, a plain CSV is no longer enough. The most common richer formats are:

- **GraphML** (`.graphml`): XML-based, very widely supported (Gephi, Cytoscape, networkx, igraph).
- **GML** (`.gml`): Simpler text format, used historically by Pajek and many classic datasets.
- **JSON** (e.g. node-link JSON): The format of choice for web visualisations.

`igraph` reads all of these out of the box via `read_graph()`.

EXERCISE:

* Open `medici.gml` and `ppi_network_prediction.graphml` in a text editor.
* Explore the code below

```{r}
#| label: c-load-gml-graphml
g_gml     <- read_graph(file.path(path_data, "medici.gml"), format = "gml")
g_graphml <- read_graph(file.path(path_data, "ppi_network_prediction.graphml"),
                        format = "graphml")

cat(sprintf("medici.gml : n = %d, m = %d\n", vcount(g_gml), ecount(g_gml)))
cat(sprintf("ppi.graphml: n = %d, m = %d\n", vcount(g_graphml), ecount(g_graphml)))
cat("GraphML vertex attributes:", vertex_attr_names(g_graphml), "\n")

# Writing to a different format is also easy
tmp_path <- tempfile(fileext = ".graphml")
write_graph(g_gml, tmp_path, format = "graphml")
```

```{r}
#| label: c-gml-dataframes
# You can see the adjacency list
head(igraph::as_data_frame(g_gml, what = "edges"))

# and the node properties
head(igraph::as_data_frame(g_gml, what = "vertices"))
```

---

## Section 2: Build real networks

### 2.1 From a data frame (Google Sheets URL)

We collected a small network during the warm-up game. The data lives in a Google Sheet, published as CSV. We can read it straight into R and convert to an `igraph` graph.

Network here: <https://tinyurl.com/network-game>

```{r}
#| label: c-game
# Load an edgelist and turn it into a graph.
# Small change to try: set `show_labels <- FALSE` if the labels overlap.
url <- "https://docs.google.com/spreadsheets/d/e/2PACX-1vSuZC86KjXYKSPr0Nw7mfRha4zwea3aMw-gTKliVcbRt_m3NRiCEyxbH_d5M8MBL0LWayg1WDmnqBET/pub?gid=0&single=true&output=csv"
show_labels <- TRUE

# read edgelist (falls back to a small edgelist if you are offline)
df <- tryCatch(
  read_csv(url, show_col_types = FALSE),
  error = function(e) tibble(P1 = c("A", "B", "C"), P2 = c("B", "C", "A"))
)
edge_columns <- c("P1", "P2")
head(df)

# convert to igraph
G_game <- graph_from_data_frame(df[, edge_columns], directed = FALSE)
cat(sprintf("n = %d, m = %d\n", vcount(G_game), ecount(G_game)))

# Visualize
set.seed(1)
plot(G_game,
     layout = layout_with_fr(G_game),
     vertex.size = 6, vertex.color = "#e6af2e",
     edge.color = "lightgray",
     vertex.label = if (show_labels) V(G_game)$name else NA,
     vertex.label.cex = 0.6, vertex.label.color = "#191716",
     main = "Network game edgelist")
```

### 2.2 Twitter network (TSV edgelist)

`ic2s2_netsci_3.tsv` is a small mention/reply/retweet network collected around the IC2S2 and NetSci conferences. It is a plain TSV with `source` and `target` columns.

```{r}
#| label: c-twitter
# Small change to try: switch `directed` to TRUE and compare n/m.
directed <- FALSE
df_tw <- read_tsv(file.path(path_data, "ic2s2_netsci_3.tsv"),
                  show_col_types = FALSE)
head(df_tw)

G_twitter <- graph_from_data_frame(df_tw[, c("source", "target")],
                                   directed = directed)
G_twitter <- simplify(G_twitter, remove.multiple = FALSE, remove.loops = TRUE)
cat(sprintf("Twitter: n = %d, m = %d\n", vcount(G_twitter), ecount(G_twitter)))

set.seed(1)
plot(G_twitter,
     layout = layout_with_fr(G_twitter),
     vertex.size = 3, vertex.color = "#3d348b",
     vertex.label = NA,
     edge.color = "lightgray", edge.width = 0.3,
     main = "Twitter network around IC2S2 NetSci")
```

### 2.3 Protein-Protein Interaction (PPI) network (GraphML)

`ppi_network_prediction.graphml` is a small yeast Protein-Protein Interaction network. Because it is already a GraphML file, we get nodes and attributes for free.

```{r}
#| label: c-ppi
# Read the network
G_ppi <- read_graph(file.path(path_data, "ppi_network_prediction.graphml"),
                    format = "graphml")
G_ppi <- as_undirected(G_ppi, mode = "collapse")
cat(sprintf("PPI: n = %d, m = %d\n", vcount(G_ppi), ecount(G_ppi)))

set.seed(1)
plot(G_ppi,
     layout = layout_with_fr(G_ppi),
     vertex.size = 2, vertex.color = "#3d348b",
     vertex.label = NA,
     edge.color = "lightgray", edge.width = 0.3,
     main = "Yeast PPI network")
```

---

> **Bring your own data**
>
> The patterns above (load a CSV -> build a graph -> visualise) work for **any** edgelist. If you brought your own data, this is a good moment to load it:
>
> 1. Save your edges as a 2-column CSV with header `source,target`.
> 2. Use the pattern from Section 2.1 to read it.
> 3. Visualise it with `plot()`.
>
> Try this with your own network before moving on.

```{r}
#| label: c-byod
# (Optional) Paste your own loading code below
# G_mine <- graph_from_data_frame(read_csv("my_data.csv"), directed = FALSE)
```

---

## Section 3: Visualizing networks

A good plot of a network can be incredibly illuminating - and a bad plot can be actively misleading. The same graph can look completely different depending on the layout algorithm.

### 3.1 The hairball problem

When a network has thousands of edges, layouts produce an unreadable blob (a "hairball"). Let's see this with the Wikipedia voting network.

```{r}
#| label: c-wiki
# Load the Wiki-Vote edgelist and plot it.
# Small change to try: change `layout_with_fr()` to `layout_randomly()`.
# Load directed Wiki-Vote network (the file starts with comment lines: #)
df_wiki <- read.table(file.path(path_data, "wiki-Vote.txt"),
                      header = FALSE, sep = "\t", comment.char = "#",
                      col.names = c("source", "target"))
G_wiki <- graph_from_data_frame(df_wiki, directed = TRUE)

# Keep only the largest weakly connected component
# (otherwise unconnected components fly away)
comps_wiki <- components(G_wiki, mode = "weak")
G_lcc <- induced_subgraph(G_wiki,
                          which(comps_wiki$membership == which.max(comps_wiki$csize)))

# Plot with a force-directed layout
set.seed(1)
plot(G_lcc,
     layout = layout_with_fr(G_lcc),
     vertex.size = 1, vertex.color = "#3d348b",
     vertex.label = NA,
     edge.color = "lightgray", edge.width = 0.05, edge.arrow.size = 0)
```

### 3.2 Visual encoding: size and colour carry information

A plain plot of a network only shows connectivity. By mapping **node size** to a centrality measure and **node colour** to a community label, we can pack much more information into the same chart.

We use the Karate club network (small enough to see every label).

```{r}
#| label: c-karate
# Karate club: built into igraph as the "Zachary" graph.
# (The Python version loads it from networkx; here we attach the club
#  factions by hand, since make_graph("Zachary") has no attributes.)
G_kar <- make_graph("Zachary")
mr_hi_members <- c(1, 2, 3, 4, 5, 6, 7, 8, 9, 11, 12, 13, 14, 17, 18, 20, 22)
V(G_kar)$club <- ifelse(seq_len(vcount(G_kar)) %in% mr_hi_members,
                        "Mr. Hi", "Officer")

cat(sprintf("Karate club: n = %d, m = %d\n", vcount(G_kar), ecount(G_kar)))

degrees <- degree(G_kar)
layout_kar <- layout_with_kk(G_kar)

# Color by real club/faction
node_colors <- ifelse(V(G_kar)$club == "Mr. Hi", palette_ns[1], palette_ns[2])

# Label only the two key nodes: 1 = Mr. Hi, 34 = Officer
node_labels <- rep("", vcount(G_kar))
node_labels[1]  <- "Mr. Hi"
node_labels[34] <- "Officer"

op <- par(mfrow = c(1, 2), mar = c(1, 1, 3, 1))

# Left: default network
plot(G_kar, layout = layout_kar, main = "Default network")

# Right: faction colors, degree sizes, leader labels
plot(G_kar, layout = layout_kar,
     vertex.color = node_colors,
     vertex.size = 5 + degrees,
     vertex.label = node_labels,
     vertex.label.cex = 1.1, vertex.label.color = "#191716",
     vertex.frame.color = NA,
     edge.color = "gray",
     main = "Karate club factions")
par(op)
```

### 3.4 Visualisation do's and don'ts

**Do**

- Think first about the **main message** you want to communicate. The other points below are contingent on that message.
- Map **node size** to importance (degree, centrality, frequency).
- Map **node colour** to a categorical attribute (community, type, label).
- Use **edge alpha** or thin edges to reduce clutter.
- Pick a **layout** that fits your message (force-directed, circular, hierarchical, ...).
- Label only the **most central** or most interesting nodes.

**Don't**

- Plot a 10-000 node hairball and call it a visualisation.
- Trust the *exact* positions of force-directed layouts - only relative positions are meaningful.

---

## Section 4: Describing networks

Once we have a graph, we want **numbers** that summarise its structure. In this section we compute the classical descriptive statistics on the PPI network and compare to the Twitter network (or your network)

```{r}
#| label: c-pick-graph
# Small change to try: set `network_choice <- "ppi"` after finishing Twitter.
network_choice <- "twitter"  # options: "ppi", "twitter"
g <- if (network_choice == "ppi") G_ppi else G_twitter
cat(sprintf("Working network (%s): n = %d, m = %d\n",
            network_choice, vcount(g), ecount(g)))
```

### 4.1 Connectedness

A *connected component* is a maximal set of nodes such that any two are reachable from each other.

```{r}
#| label: c-components
# Only function to notice: `components()`.
comps <- components(g, mode = "weak")
component_sizes <- sort(comps$csize, decreasing = TRUE)
cat(sprintf("Number of components: %d\n", comps$no))
cat("Sizes of the 20 largest components:", head(component_sizes, 20), "\n")

g_lcc <- induced_subgraph(g, which(comps$membership == which.max(comps$csize)))
cat(sprintf("Largest component: n = %d, m = %d\n", vcount(g_lcc), ecount(g_lcc)))
```

### 4.2 Diameter and average shortest path

- **Diameter**: the longest shortest path between any pair of nodes.
- **Average shortest path**: the mean over all pairs.

Many real-world networks have *very* short paths between nodes ("small world").

```{r}
#| label: c-diameter
cat(sprintf("Diameter            : %d\n", diameter(g_lcc)))
cat(sprintf("Average path length : %.3f\n", mean_distance(g_lcc)))
```

### 4.3 Density

Density = (edges present) / (edges possible). Most real networks are extremely sparse.

```{r}
#| label: c-density
cat(sprintf("Density: %.5f\n", edge_density(g_lcc)))
```

### 4.4 Transitivity and average clustering

- **Global transitivity** = 3 x (#triangles) / (#connected triples). The "ratio of closed triangles".
- **Average local clustering** = mean local clustering coefficient over nodes.

The two can differ a lot in heterogeneous networks.

```{r}
#| label: c-transitivity
# Two ways to summarise triangle closure.
# The arguments to notice are `type = "global"` and `type = "average"`.
cat(sprintf("Global transitivity        : %.4f\n",
            transitivity(g_lcc, type = "global")))
cat(sprintf("Average local clustering   : %.4f\n",
            transitivity(g_lcc, type = "average", isolates = "zero")))
```

### 4.5 Degree assortativity

Are high-degree nodes connected to other high-degree nodes (assortative, > 0) or to low-degree ones (disassortative, < 0)?

```{r}
#| label: c-assortativity
# Degree assortativity asks whether high-degree nodes connect to high-degree nodes.
cat(sprintf("Degree assortativity: %.4f\n", assortativity_degree(g_lcc)))
```

### 4.6 Robustness to failures vs. fragility to targeted attacks

We remove nodes one at a time, either:

- **targeted**: highest-degree node first (think: a smart adversary),
- **random**: uniformly at random (think: random failures).

After each removal we track the share of nodes still in the giant component.

```{r}
#| label: c-attack
# Small changes to try: lower/raise `n_remove`.
attack_network <- function(g_in, how = "targeted", n_remove = 200, seed = 0) {
  set.seed(seed)
  g2 <- g_in
  shares <- numeric(n_remove)
  for (i in seq_len(n_remove)) {
    if (vcount(g2) <= 1) {
      shares[i] <- 0
      next
    }
    node_idx <- if (how == "targeted") which.max(degree(g2))
                else sample.int(vcount(g2), 1)
    g2 <- delete_vertices(g2, node_idx)
    cc <- components(g2, mode = "weak")
    biggest <- if (cc$no > 0) max(cc$csize) else 0
    shares[i] <- 100 * biggest / max(vcount(g2), 1)
  }
  shares
}

n_remove <- if (network_choice == "ppi") 200 else 1000
# Random network (Erdos Renyi = connections are created randomly), as baseline
set.seed(1)
G_er <- sample_gnm(n = vcount(g), m = ecount(g), directed = FALSE, loops = FALSE)
er_comps <- components(G_er, mode = "weak")
G_er <- induced_subgraph(G_er, which(er_comps$membership == which.max(er_comps$csize)))

# Attack the network and the baseline
obs_targeted <- attack_network(g,    how = "targeted", n_remove = n_remove, seed = 1)
obs_random   <- attack_network(g,    how = "random",   n_remove = n_remove, seed = 2)
er_targeted  <- attack_network(G_er, how = "targeted", n_remove = n_remove, seed = 3)
er_random    <- attack_network(G_er, how = "random",   n_remove = n_remove, seed = 4)

# Plot (tidy style: long data frame + facets)
attack_df <- tibble(
  removed  = rep(100 * (seq_len(n_remove) - 1) / vcount(g), 4),
  share    = c(obs_targeted, er_targeted, obs_random, er_random),
  graph    = rep(rep(c("Observed network", "ER random graph"), each = n_remove), 2),
  scenario = rep(c("Targeted attack", "Random failures"), each = 2 * n_remove)
)

ggplot(attack_df, aes(removed, share, color = graph)) +
  geom_line(linewidth = 1) +
  facet_wrap(~ scenario) +
  scale_color_manual(values = palette_ns[1:2]) +
  ylim(0, 100) +
  labs(x = "Share of nodes removed (%)",
       y = "Share in giant component (%)",
       color = NULL)
```

### Comparison: PPI vs Twitter

| Property | Twitter | PPI |
|---------|---------|-----|
| Connectedness | 1 component, 1554 nodes | ~106 components, 809 in largest |
| Diameter | 6 (avg path 2.5) | 15 (avg path 5.6) |
| Density | 4.4% | 0.35% |
| Transitivity | 21% | 1.4% |
| Avg. clustering | 35% | 3.5% |
| Degree assortativity | 12.6% | -16.4% |
| Robustness | fairly robust, breaks around 50% | fragile, breaks at ~5% |

The Twitter network is much denser and much more clustered.

---

## Section 5: Distributions

Summary statistics throw away a lot of information. Distributions over nodes reveal heterogeneity, hubs, and heavy tails. We work with the Florentine families network here because it is small enough to plot every node.

```{r}
#| label: c-flo-setup
# Reuse the small Florentine network for node-level distributions.
# Padgett's marriage network ships with the `network` package as an adjacency
# matrix; we drop the isolated Pucci family, like networkx's
# florentine_families_graph() in the Python version.
data(flo, package = "network")
G_flo <- graph_from_adjacency_matrix(flo, mode = "undirected")
G_flo <- delete_vertices(G_flo, degree(G_flo) == 0)

g_flo <- G_flo
cat(sprintf("Florentine: n = %d, m = %d\n", vcount(g_flo), ecount(g_flo)))
cat("Families:", V(g_flo)$name, "\n")
```

### 5.1 Degree distribution

`degree(g)` returns a vector with one degree per node.

```{r}
#| label: c-degree-dist
# Compare a histogram with a CCDF for degree.
# Small change to try: set `ccdf_scale <- "linear"`.
plot_ccdf <- function(values, xlabel = "Degree d", ylabel = "p(Degree > d)",
                      scale = "log") {
  df <- tibble(d = values) |>
    count(d) |>
    arrange(d) |>
    mutate(ccdf = 1 - cumsum(n) / sum(n))
  p <- ggplot(df, aes(d, ccdf)) +
    geom_point(color = palette_ns[1]) +
    geom_line(color = palette_ns[1]) +
    labs(x = xlabel, y = ylabel)
  if (scale == "log") p <- p + scale_x_log10() + scale_y_log10()
  p
}

ccdf_scale <- "log"
degree_values <- degree(g_flo)

p_hist <- ggplot(tibble(d = degree_values), aes(d)) +
  geom_histogram(binwidth = 1, fill = palette_ns[1]) +
  labs(x = "Degree d", y = "# nodes", title = "Degree histogram")
p_ccdf <- plot_ccdf(degree_values, scale = ccdf_scale) +
  labs(title = "Degree CCDF")

p_hist
p_ccdf
```

### 5.2 Triangles per node

`igraph` has a direct `count_triangles()` function, but we can also compute it cheaply from the adjacency matrix:

> The number of triangles a node `i` participates in is `(A^3)[i, i] / 2`.

```{r}
#| label: c-triangles-per-node
# Triangles through a node come from the diagonal of A^3.
A_flo <- as_adjacency_matrix(g_flo, sparse = FALSE)
A3 <- A_flo %*% A_flo %*% A_flo
triangles_per_node <- as.integer(diag(A3) / 2)

tibble(
  family    = V(g_flo)$name,
  triangles = triangles_per_node
) |>
  arrange(desc(triangles))
```

### 5.3 Local clustering coefficient

`transitivity(g, type = "local")` returns one local clustering value per node.

```{r}
#| label: c-local-clustering
# Local clustering is one value per node.
local_clust <- transitivity(g_flo, type = "local", isolates = "zero")
clust_df <- tibble(family = V(g_flo)$name, local_clustering = local_clust)
clust_df |> arrange(desc(local_clustering))

ggplot(clust_df, aes(local_clustering)) +
  geom_histogram(bins = 10, fill = palette_ns[1]) +
  labs(x = "Local clustering coefficient", y = "# nodes",
       title = "Local clustering distribution")
```

### 5.4 Shortest path lengths

`distances(g)` returns a full pairwise distance matrix. For large networks this is slow / memory-hungry.

```{r}
#| label: c-shortest-paths
# Shortest paths are pairwise distances on the graph.
dist_matrix <- distances(g_flo)
mask <- is.finite(dist_matrix) & dist_matrix > 0
all_dists <- dist_matrix[mask]
cat(sprintf("Mean shortest path: %.3f\n", mean(all_dists)))
cat(sprintf("Diameter (max)    : %d\n", as.integer(max(all_dists))))

ggplot(tibble(d = all_dists), aes(d)) +
  geom_histogram(binwidth = 1, fill = palette_ns[1]) +
  labs(x = "Shortest path length", y = "# pairs",
       title = "Distribution of shortest paths")
```

### 5.5 PDF, CDF and CCDF on log-log scale

A power-law degree distribution looks like a straight line on a log-log CCDF plot. Let's compare the degree distributions of the Wiki, Twitter and PPI networks.

```{r}
#| label: c-distributions
# PDF/CDF/CCDF show different views of the same degree sequence.
# Small change to try: change `which_graph` to "wiki" or "ppi".
plot_distributions <- function(d, title = "", scale = "log") {
  df <- tibble(d = d) |>
    count(d) |>
    arrange(d) |>
    mutate(
      pdf  = n / sum(n),
      cdf  = cumsum(pdf),
      ccdf = 1 - cdf
    ) |>
    pivot_longer(c(pdf, cdf, ccdf), names_to = "stat", values_to = "p") |>
    mutate(stat = factor(toupper(stat), levels = c("PDF", "CDF", "CCDF")))

  p <- ggplot(df, aes(d, p)) +
    geom_point(color = palette_ns[1], size = 0.8) +
    facet_wrap(~ stat, nrow = 1) +
    labs(x = "Degree d", y = NULL, title = title)
  if (scale == "log") p <- p + scale_x_log10() + scale_y_log10()
  p
}

which_graph <- "twitter"  # try: "wiki", "twitter", "ppi"
graphs <- list(wiki = G_wiki, twitter = G_twitter, ppi = G_ppi)
plot_distributions(degree(graphs[[which_graph]]),
                   title = tools::toTitleCase(which_graph), scale = "log")
```

---

## Section 6: Reference models and permutation test

When we measure clustering = 0.19 on the Florentine network, the natural question is:

> **Is 0.19 high? Compared to what?**

The standard answer is to compare against a *reference model*: a random graph that preserves *some* property of the observed network (number of nodes, number of edges, degree sequence) but is otherwise random.

### 6.1 Three reference models

| Model | Preserves | igraph constructor |
|-------|-----------|--------------------|
| Erdos-Renyi (ER, `G(n, m)`) | n, m | `sample_gnm(n, m)` |
| Configuration model | full degree sequence | `sample_degseq(deg, method = "vl")` |
| Barabasi-Albert | preferential-attachment growth | `sample_pa(n, m, directed = FALSE)` |

### 6.2 Visualise three reference graphs for the Florentine network

```{r}
#| label: c-refmodels-plot
# Start with the Florentine network
n_flo <- vcount(g_flo)
m_flo <- ecount(g_flo)
deg_seq <- degree(g_flo)

# Create baselines
G_er  <- sample_gnm(n = n_flo, m = m_flo, directed = FALSE, loops = FALSE)
G_cfg <- sample_degseq(deg_seq, method = "vl")
G_ba  <- sample_pa(n = n_flo, m = max(1, m_flo %/% n_flo), directed = FALSE)

# Plot them
set.seed(1)  # so the layout doesn't change if the network didn't change
op <- par(mfrow = c(1, 4), mar = c(1, 1, 3, 1))
for (item in list(list("Observed", g_flo), list("Erdos-Renyi", G_er),
                  list("Configuration", G_cfg), list("Barabasi-Albert", G_ba))) {
  gx <- item[[2]]
  plot(gx,
       layout = layout_with_fr(gx),
       vertex.size = 8, vertex.color = "#3d348b",
       vertex.label = NA,
       edge.color = "lightgray",
       main = sprintf("%s\nn=%d, m=%d", item[[1]], vcount(gx), ecount(gx)))
}
par(op)
```

### 6.3 Confidence-interval comparison

We generate many random graphs of the same size, compute a metric on each, and compare the observed value to the 5%-95% percentile range.

```{r}
#| label: c-confint
conf_int_ig <- function(g_obs, fun, iters = 100, model = "er", seed = 0) {
  set.seed(seed)
  n <- vcount(g_obs)
  m <- ecount(g_obs)
  deg_seq <- degree(g_obs)
  vals <- numeric(iters)
  for (i in seq_len(iters)) {
    Gr <- switch(model,
      er     = sample_gnm(n, m, directed = FALSE, loops = FALSE),
      config = sample_degseq(deg_seq, method = "vl"),
      ba     = sample_pa(n, max(1, m %/% n), directed = FALSE)
    )
    vals[i] <- fun(Gr)
  }
  quantile(vals[is.finite(vals)], probs = c(0.05, 0.95))
}

# small change: use assortativity_degree() instead
observed <- transitivity(g_flo, type = "global")
metric_fun <- function(G) transitivity(G, type = "global")

cat(sprintf("Observed: %.3f\n", observed))
for (model in c("er", "config", "ba")) {
  ci <- conf_int_ig(g_flo, metric_fun, iters = 100, model = model, seed = 1)
  cat(sprintf("  %-7s: [%.3f, %.3f]\n", model, ci[1], ci[2]))
}
```

### 6.4 Permutation test for attribute assortativity

The Florentine network can be annotated with a political faction: each family was either a **Medici** ally (`m`) or an **oligarch** (`o`). Do allied families marry each other more than chance would predict?

We test this by permuting the labels many times and comparing the observed assortativity to the null distribution.

```{r}
#| label: c-permtest-setup
# Attach a simple political faction label to each family.
# Small change to try: change one family from "o" to "m" and rerun the test below.
ns_labels <- c("Acciaiuoli", "Medici", "Castellani", "Peruzzi", "Strozzi",
               "Barbadori", "Ridolfi", "Tornabuoni", "Albizzi", "Salviati",
               "Pazzi", "Bischeri", "Guadagni", "Ginori", "Lamberteschi")
classes <- c("m", "m", "o", "o", "o", "o", "o", "m", "m", "m", "o", "o", "o", "o", "o")
name_to_class <- setNames(classes, ns_labels)
V(g_flo)$faction <- coalesce(name_to_class[V(g_flo)$name], "o")

faction_int <- ifelse(V(g_flo)$faction == "m", 0L, 1L)
node_color  <- ifelse(V(g_flo)$faction == "m", "gray", "cornflowerblue")
obs_assort  <- assortativity_nominal(g_flo, types = faction_int + 1, directed = FALSE)
cat(sprintf("Observed attribute assortativity: %.4f\n", obs_assort))

set.seed(0)
plot(g_flo,
     layout = layout_with_fr(g_flo),
     vertex.color = node_color,
     vertex.size = 14,
     vertex.frame.color = NA,
     vertex.label = V(g_flo)$name, vertex.label.cex = 0.7,
     vertex.label.color = "#191716",
     edge.color = "gray")
```

```{r}
#| label: c-permtest-run
# Permutation test for "is this assortativity surprising?"
# Small change to try: set `iters <- 10000` for a smoother histogram.
set.seed(0)
iters <- 1000
null_values <- numeric(iters)
for (i in seq_len(iters)) {
  permuted <- sample(faction_int)
  null_values[i] <- assortativity_nominal(g_flo, types = permuted + 1,
                                          directed = FALSE)
}

ggplot(tibble(x = null_values), aes(x)) +
  geom_histogram(binwidth = 0.05, fill = palette_ns[4], color = "white") +
  geom_vline(xintercept = obs_assort, linetype = "dashed",
             color = palette_ns[1], linewidth = 1) +
  annotate("text", x = obs_assort, y = Inf, hjust = -0.1, vjust = 2,
           label = sprintf("observed = %.3f", obs_assort),
           color = palette_ns[1]) +
  labs(x = "Attribute assortativity under random labels",
       y = "# permutations")

p_value <- mean(null_values >= obs_assort)
cat(sprintf("One-sided p-value (null >= observed): %.4f\n", p_value))
```

**Interpretation.** If the p-value is small (say < 0.05) we reject the null hypothesis that faction labels are independent of the marriage structure: families *did* preferentially marry within their political faction.

---

That's the end of practical 1a. You now have a working toolbox to:

1. Load a network from any common format.
2. Visualise it sensibly.
3. Describe it with numbers.
4. Plot its key distributions.
5. Test whether what you observe is "surprising" against a reasonable null.
