---
title: "1b. Matrix Multiplication and Centrality (R / igraph edition)"
format:
  html:
    toc: true
    toc-depth: 2
    code-fold: false
    embed-resources: true
    html-math-method: katex
execute:
  warning: false
  message: false
---

# 1b. Matrix Multiplication and Centrality

Welcome to the afternoon practical! In this session we connect **linear algebra** to **networks**: the adjacency matrix is not just storage, it is a tool that answers questions about the network when you multiply it by things.

**Learning objectives**

- Represent a network as an adjacency matrix and convert back and forth between formats.
- Understand what `A %*% x` and `A %*% A` answer about the network.
- Use `A^n` to count walks of length `n`, triangles, and the local clustering coefficient.
- Compute and compare the main centrality measures (degree, closeness, betweenness, eigenvector, PageRank, harmonic, HITS).

**Sections**

0. Setup and helpers
1. Network representations
2. Matrix multiplication = combining neighbour attributes
3. Matrix multiplication and paths
4. Node centrality on the Florentine network
5. Centrality on real networks (Twitter, PPI)
6. Eigenvector and PageRank via the power method
7. Centrality side-by-side comparison

## Section 0: Setup and helpers

```{r}
#| label: c-install
# If you are missing packages, install them by uncommenting:
# install.packages(c("igraph", "dplyr", "readr", "tidyr", "ggplot2", "RColorBrewer", "network"))
```

```{r}
#| label: c-path
# Path to the data folder (relative to this .qmd file)
path_data <- "../../../Data/"
```

```{r}
#| label: c-setup
# Core libraries
library(igraph)        # Primary network library for this practical
library(dplyr)         # Data manipulation
library(readr)         # Reading CSV / TSV files
library(tidyr)         # Reshaping data
library(ggplot2)       # Plotting
library(RColorBrewer)  # Colour palettes

# Visual style for the whole notebook
palette_ns <- c("#3d348b", "#e6af2e", "#191716", "#e0e2db")
theme_set(
  theme_minimal(base_size = 13) +
    theme(
      panel.grid.minor = element_blank(),
      panel.grid.major = element_line(color = "lightgray")
    )
)

# A diverging palette comparable to matplotlib's "coolwarm"
coolwarm_palette <- colorRampPalette(rev(brewer.pal(11, "RdBu")))
value_to_color <- function(values, n = 256) {
  vals <- as.numeric(values)
  rng <- range(vals, finite = TRUE)
  if (diff(rng) < 1e-12) rng <- rng + c(-1, 1) * 1e-9
  idx <- pmax(1, pmin(n, round(1 + (n - 1) * (vals - rng[1]) / diff(rng))))
  coolwarm_palette(n)[idx]
}

# Reproducibility seed (igraph layouts use R's RNG under the hood)
set.seed(1)
```

```{r}
#| label: c-helpers
# --- Helper functions for visualisation and conversion (igraph edition) ---
# These mirror the Python helpers; they use only igraph + base R graphics.

plot_network <- function(g, values = NULL, with_labels = TRUE,
                         main = "", layout = NULL) {
  # Draw an igraph network. Node size and colour are scaled to `values`.
  if (is.null(values)) {
    deg <- degree(g)
    values <- deg / max(1, max(deg))
  }
  values <- as.numeric(values)

  node_colors <- value_to_color(values)
  span <- max(values) - min(values)
  if (span < 1e-12) span <- 1
  sizes <- 12 + 25 * (values - min(values)) / span

  labels <- if (with_labels && "name" %in% vertex_attr_names(g)) V(g)$name else NA

  if (is.null(layout)) {
    set.seed(1)
    layout <- layout_with_fr(g)
  }
  plot(g, layout = layout,
       vertex.size = sizes, vertex.color = node_colors,
       vertex.label = labels, vertex.label.cex = 0.8,
       vertex.label.color = "#191716",
       edge.color = "darkgray",
       main = main)
}

plot_table <- function(A, names_v) {
  # Render an adjacency matrix as a grid of numbers, highlighting non-zero cells.
  n <- nrow(A)
  plot(NULL, xlim = c(0.5, n + 0.5), ylim = c(0.5, n + 0.5),
       axes = FALSE, xlab = "", ylab = "", asp = 1)
  for (i in seq_len(n)) {
    for (j in seq_len(n)) {
      bg <- if (A[i, j] > 0) "#e6af2e" else "white"
      rect(j - 0.5, n - i + 0.5, j + 0.5, n - i + 1.5,
           col = bg, border = "lightgray")
      text(j, n - i + 1, A[i, j], cex = 0.8)
    }
  }
  axis(3, at = seq_len(n), labels = names_v, las = 2, cex.axis = 0.7,
       tick = FALSE, line = -1)
  axis(2, at = rev(seq_len(n)), labels = names_v, las = 1, cex.axis = 0.7,
       tick = FALSE, line = -1)
}

plot_network_adj <- function(g, values = NULL, main = "") {
  # Three-panel display: network, adjacency-as-table, adjacency spy plot.
  op <- par(mfrow = c(1, 3), mar = c(2, 3, 3, 1))
  plot_network(g, values = values, main = main)

  A <- as_adjacency_matrix(g, sparse = FALSE)
  names_v <- if ("name" %in% vertex_attr_names(g)) {
    V(g)$name
  } else {
    as.character(seq_len(vcount(g)))
  }
  plot_table(A, names_v)

  # spy plot: just the non-zero positions
  nz <- which(A != 0, arr.ind = TRUE)
  plot(nz[, "col"], nrow(A) - nz[, "row"] + 1,
       pch = 15, cex = 2, col = "#3d348b",
       xlim = c(0.5, ncol(A) + 0.5), ylim = c(0.5, nrow(A) + 0.5),
       axes = FALSE, xlab = "", ylab = "", asp = 1)
  axis(3, at = seq_len(ncol(A)), labels = names_v, las = 2, cex.axis = 0.7,
       tick = FALSE, line = -1)
  axis(2, at = rev(seq_len(nrow(A))), labels = names_v, las = 1, cex.axis = 0.7,
       tick = FALSE, line = -1)
  par(op)
}

adj_to_net <- function(A, names_v = c("Alice", "Bob", "John", "Amy", "Mike", "Rose")) {
  # Build a directed igraph from a (dense) adjacency matrix and attach names.
  A <- as.matrix(A)
  g <- graph_from_adjacency_matrix((A > 0) * 1, mode = "directed")
  if (!is.null(names_v) && length(names_v) == vcount(g)) {
    V(g)$name <- names_v
  }
  g
}
```

---

## Section 1: Network representations

We will work with the same tiny 6-person network in two flavours, directed and undirected, so we can see how the adjacency matrix changes.

### 1.1 Build the toy directed and undirected networks

Edges (read as "A points to B" for the directed version):

```
Alice -> Bob
John  -> Alice
Bob   -> John
Amy   -> John
Mike  -> John
Rose  -> Alice
Mike  -> Rose
```

```{r}
#| label: c-toy-build
# Create small network
edges <- tibble(
  from = c("Alice", "John",  "Bob",  "Amy",  "Mike", "Rose",  "Mike"),
  to   = c("Bob",   "Alice", "John", "John", "John", "Alice", "Rose")
)

G_dir   <- graph_from_data_frame(edges, directed = TRUE)
G_undir <- graph_from_data_frame(edges, directed = FALSE)

plot_network_adj(G_dir, main = "Directed")
plot_network_adj(G_undir, main = "Undirected (matrix is symmetric)")
```

Notice that the undirected adjacency matrix is **symmetric** (`A == t(A)`), while the directed one is generally not.

### 1.2 Convert between formats

The three common representations are:

1. **Edgelist** (data frame of source/target) — easy to read, easy to import/export.
2. **Adjacency matrix** (`matrix`) — perfect for linear algebra.
3. **Graph object** — what `igraph` actually uses to compute on.

A good network library lets you round-trip between all three. Let's do it.

```{r}
#| label: c-convert
# To edgelist (vertex ids)
df <- as_edgelist(G_dir, names = FALSE) |>
  as.data.frame() |>
  setNames(c("source", "target"))
df

# get the real names
vertex_names <- V(G_dir)$name
df$source <- vertex_names[df$source]
df$target <- vertex_names[df$target]
df

# To adjacency (as a matrix with named rows/columns to visualize it)
A <- as_adjacency_matrix(G_dir, sparse = FALSE)
A
```

### 1.3 Add node metadata

We attach a `children` attribute to each person. This will become the vector we multiply the adjacency matrix by in the next section.

```{r}
#| label: c-metadata
# Attach one node attribute in graph vertex order.
cat("Node order:", V(G_dir)$name, "\n")
children_map <- c(Alice = 2, Bob = 0, John = 2, Amy = 0, Mike = 1, Rose = 5)
V(G_dir)$children   <- children_map[V(G_dir)$name]
V(G_undir)$children <- children_map[V(G_undir)$name]

tibble(person = V(G_dir)$name, children = V(G_dir)$children)
```

```{r}
#| label: c-attributes
# Print node and edge attributes
print(vertex_attr_names(G_dir))
print(edge_attr_names(G_dir))
```

---

## Section 2: Matrix multiplication = combining neighbour attributes

> **The big idea.** Multiplying by the adjacency matrix `A` is the network's way of **asking a question about neighbours**.
>
> - `A %*% x` asks: *"for each node, what is the sum of my neighbours' values of `x`?"*
> - `A %*% A` (= `A^2`) asks: *"for each pair of nodes, how many length-2 paths connect them?"*
> - `A %*% A %*% A` (= `A^3`) asks the same about length-3 paths, and its diagonal counts triangles.
>
> Every linear-algebra operation we do in this practical is a small variation on those two questions.

### 2.1 Average number of children of your friends

For the **undirected** version of the network we want, for each node:

`average_children_of_neighbours = (sum of neighbours' children) / (number of neighbours)`

which in matrix form is:

`avg = (A %*% children) / degree`

```{r}
#| label: c-avg-children-setup
# Create adjacency matrix
A <- as_adjacency_matrix(G_undir, sparse = FALSE)
# And the number of children
x <- as.numeric(V(G_undir)$children)

# Plot network (node colour: number of own children)
plot_network_adj(G_undir, values = x,
                 main = "Node colour: own children")
```

```{r}
#| label: c-avg-children
# Calculate sum
sum_children <- as.numeric(A %*% x)
# Divide by degree
degree_u <- rowSums(A)
avg_children <- sum_children / degree_u

# Put into a data frame to explore and visualize
result <- tibble(
  person           = V(G_undir)$name,
  own_children     = as.integer(x),
  neighbours_total = as.integer(sum_children),
  degree           = as.integer(degree_u),
  avg_neighbour    = round(avg_children, 2)
)
result
```

### 2.2 What does `t(A) %*% x` mean for a directed network?

For the directed `G_dir`, `A[i, j] = 1` means **i points to j**. So `(A %*% x)[i]` sums values of nodes that `i` points to (its *out-neighbours*). What does `(t(A) %*% x)[i]` represent?

```{r}
#| label: c-transpose
# Transpose the matrix to reverse the direction of the question.
A <- as_adjacency_matrix(G_dir, sparse = FALSE)
x <- as.numeric(V(G_dir)$children)
plot_network_adj(G_dir, values = x)

out_neighbour_sum <- as.numeric(A %*% x)
in_neighbour_sum  <- as.numeric(t(A) %*% x)

tibble(
  person = V(G_dir)$name,
  `A x (out-neighbours)`    = as.integer(out_neighbour_sum),
  `t(A) x (in-neighbours)`  = as.integer(in_neighbour_sum)
)
# t(A) %*% x is the network asking: "for each node, what is the sum of values
# coming from my IN-neighbours (the nodes that point at me)?"
```

---

## Section 3: Matrix multiplication and paths

`A^n` counts **walks** of length `n`: `(A^n)[i, j]` is the number of length-`n` walks from `i` to `j`. This single fact powers:

- counting nodes reachable in at most `k` steps,
- counting triangles,
- the local clustering coefficient.

### 3.1 Walks of length 1, 2 and 3

We compute `A`, `A^2`, `A^3` on the directed network.

`(A + A^2 + A^3) > 0` is the "can I reach node `j` from `i` in 1, 2 or 3 steps?" matrix.

```{r}
#| label: c-walks-plot
plot_network(G_dir)
```

```{r}
#| label: c-walks
# Powers of A count walks of fixed length.
# Combine paths of length 1, 2 and 3, then count reachable nodes.
A <- as_adjacency_matrix(G_dir, sparse = FALSE)
M <- (A + A %*% A + A %*% A %*% A) > 0  # count reachable nodes once
diag(M) <- FALSE                        # subtract yourself
reachable_le_3 <- rowSums(M)            # sum all nodes reachable

tibble(person = V(G_dir)$name, `reachable within 3 steps` = reachable_le_3)
```

### 3.2 Triangles

The diagonal of `A^3` counts closed walks of length 3 — i.e. triangles, with multiplicity. For an **undirected** simple graph each triangle is counted **6 times** in `trace(A^3)` (3 vertices x 2 directions), so the total number of triangles is `trace(A^3) / 6`.

Compare with `cliques(g, min = 3, max = 3)`.

```{r}
#| label: c-triangles
# For undirected graphs, trace(A^3) / 6 counts triangles.
A <- as_adjacency_matrix(G_undir, sparse = FALSE)
n_tri_matrix <- as.integer(sum(diag(A %*% A %*% A)) / 6)
n_tri_igraph <- length(cliques(G_undir, min = 3, max = 3))
cat(sprintf("Triangles via matrix : %d\n", n_tri_matrix))
cat(sprintf("Triangles via igraph : %d\n", n_tri_igraph))
```

### 3.3 Local clustering coefficient

The local clustering coefficient of node `i` is

```
C_i = (triangles through i) / (number of pairs of neighbours of i)
    = (A^3)[i, i] / 2     /     ( k_i * (k_i - 1) / 2 )
```

Compare to `transitivity(g, type = "local")`.

```{r}
#| label: c-clustering
# Local clustering compares actual triangles to possible triangles.
A <- as_adjacency_matrix(G_undir, sparse = FALSE)
tri <- diag(A %*% A %*% A) / 2
degree_u <- rowSums(A)
potential <- degree_u * (degree_u - 1) / 2
C_manual <- tri / potential
C_igraph <- transitivity(G_undir, type = "local", isolates = "zero")

tibble(
  node          = V(G_undir)$name,
  triangles     = tri,
  potential     = potential,
  `C (manual)`  = round(C_manual, 3),
  `C (igraph)`  = round(C_igraph, 3)
)
```

---

## Section 4: Node centrality

Centrality measures try to capture, in a single number, **how important** a node is. They disagree because they answer different questions:

| Measure          | Question it answers                                                    |
|------------------|------------------------------------------------------------------------|
| **Degree**       | How many direct connections do I have?                                 |
| **Closeness**    | How short is my average distance to everyone else?                     |
| **Betweenness**  | How often do I sit on a shortest path between two other nodes?         |
| **Eigenvector**  | How well-connected are my neighbours (recursively)?                    |
| **PageRank**     | Like eigenvector, but for directed networks and with a damping factor. |
| **Harmonic**     | Closeness that gracefully handles disconnected components.             |

### 4.1 Florentine families network

We load the classic Florentine marriage network. NetworkX ships it as a convenience constructor; in R it comes with the `network` package (Padgett's data) as an adjacency matrix.

```{r}
#| label: c-flo-load
# Load the Florentine families network (drop the isolated Pucci family,
# like networkx's florentine_families_graph()).
data(flo, package = "network")
G_flo <- graph_from_adjacency_matrix(flo, mode = "undirected")
G_flo <- delete_vertices(G_flo, degree(G_flo) == 0)
cat(sprintf("Florentine: n = %d, m = %d\n", vcount(G_flo), ecount(G_flo)))
cat("Families:", V(G_flo)$name, "\n")
```

### 4.2 Compute the five centralities

We collect them in a data frame for inspection.

```{r}
#| label: c-flo-centralities
# Compute several centralities and compare rankings.
cent <- tibble(
  family      = V(G_flo)$name,
  degree      = degree(G_flo),
  closeness   = closeness(G_flo),
  betweenness = betweenness(G_flo),
  eigenvector = eigen_centrality(G_flo)$vector,
  pagerank    = page_rank(G_flo)$vector,
  harmonic    = harmonic_centrality(G_flo)
)

sort_by <- "betweenness"  # try: "degree", "closeness", "eigenvector", "pagerank"
cent |>
  arrange(desc(.data[[sort_by]])) |>
  mutate(across(where(is.numeric), \(v) round(v, 3)))
```

### 4.3 Visualise each centrality on the network

The node size and colour both encode the centrality value, so you can scan the picture and spot the "winners".

```{r}
#| label: c-flo-plot
# Small change to try: set `centrality_to_plot <- "degree"`.
centrality_to_plot <- "betweenness"
vals <- cent[[centrality_to_plot]]
plot_network(G_flo, values = vals,
             main = paste("Florentine families -", centrality_to_plot, "centrality"))
```

---

## Section 5: Centrality on real networks

### 5.1 Twitter network (`ic2s2_netsci_3.tsv`)

Load the directed Twitter network and report the top nodes by closeness, betweenness, PageRank and hub/authority score (HITS).

```{r}
#| label: c-twitter-cent
df <- read_tsv(file.path(path_data, "ic2s2_netsci_3.tsv"), show_col_types = FALSE)
G_tw <- graph_from_data_frame(df[, c("source", "target")], directed = TRUE)
G_tw <- simplify(G_tw, remove.multiple = FALSE, remove.loops = TRUE)
cat(sprintf("Twitter network: n = %d, m = %d\n", vcount(G_tw), ecount(G_tw)))

names_tw <- V(G_tw)$name

top_n_tbl <- function(values, names_v, label, n = 15) {
  tibble(name = names_v, !!label := values) |>
    slice_max(.data[[label]], n = n)
}

hits <- hits_scores(G_tw)
cc <- top_n_tbl(closeness(G_tw, mode = "all"),           names_tw, "closeness")
bc <- top_n_tbl(betweenness(G_tw, directed = TRUE),      names_tw, "betweenness")
pr <- top_n_tbl(page_rank(G_tw, directed = TRUE)$vector, names_tw, "pagerank")
hb <- top_n_tbl(hits$hub,                                names_tw, "hub_score")
ab <- top_n_tbl(hits$authority,                          names_tw, "authority_score")

list(cc, bc, pr, hb, ab) |>
  Reduce(\(a, b) full_join(a, b, by = "name"), x = _) |>
  arrange(desc(authority_score))
```

### 5.2 Protein-protein interaction network

Same exercise on the yeast PPI network. The top nodes by closeness/betweenness/eigenvector tend to be hub proteins central to cellular machinery.

> `YLR291C` is a known hub: it is involved in regulation of translational initiation.

```{r}
#| label: c-ppi-cent
G_ppi <- read_graph(file.path(path_data, "ppi_network_prediction.graphml"),
                    format = "graphml")
G_ppi <- as_undirected(G_ppi, mode = "collapse")
# Use the "label" attribute as the node name if available
if ("label" %in% vertex_attr_names(G_ppi)) V(G_ppi)$name <- V(G_ppi)$label
cat(sprintf("PPI: n = %d, m = %d\n", vcount(G_ppi), ecount(G_ppi)))

names_ppi <- V(G_ppi)$name

cc_p <- top_n_tbl(degree(G_ppi),                    names_ppi, "degree")
bc_p <- top_n_tbl(betweenness(G_ppi),               names_ppi, "betweenness")
ec_p <- top_n_tbl(eigen_centrality(G_ppi)$vector,   names_ppi, "eigenvector")

list(cc_p, bc_p, ec_p) |>
  Reduce(\(a, b) full_join(a, b, by = "name"), x = _)
```

### 5.3 Discussion

Look at the rankings above and discuss with your neighbour, in the Twitter network:

- **Who would you choose to spread news quickly** (assuming the news is *always* passed on)?
- **Who connects different communities** (acts as a bridge between groups)?
- **Who would be best for spreading unreliable news** (assuming the news is sometimes *not* passed on)?

There are no single right answers. The answers depend on which centrality you trust for each question, and that is exactly the point of having more than one.

---

## Section 6: Eigenvector centrality and PageRank, by hand

[advanced, only if you love math]

### Scaffold A — the idea

Eigenvector centrality is built on a simple, almost circular, idea:

> **A node is important if its neighbours are important.**

That sounds like we're chasing our own tail, but watch what happens when we turn it into an equation.

### Scaffold B — turning the idea into a formula

Call each node's importance `x_i`. The statement *"my importance is the sum of my neighbours' importances"* reads:

$$ x_i = \sum_j{A_{i, j}  x_j} $$

In vector form, that's `x = A x`. Almost an eigenvector equation, but `A x` will keep growing (or shrinking) every time we apply `A`, so to keep magnitudes stable we normalise:

`x = (A x) / λ`

where λ is just a normalising constant. Rearranging:

`A x = λ x`

That **is** the eigenvalue equation. The eigenvector of `A` with the largest eigenvalue λ is exactly what we call **eigenvector centrality**.

### Scaffold C — the power method

You don't need to call an eigensolver to find `x`. Start with `x = (1, 1, …, 1)` and repeat:

```
x  <-  A %*% x
x  <-  x / norm(x)
```

After enough iterations `x` converges to the dominant eigenvector. That repeated multiply-and-normalise is the **power method**.

### 6.1 Eigenvector centrality with the power method

We run the power method on `G_undir` and print the result side-by-side with `eigen_centrality()`.

```{r}
#| label: c-power-eig
# Watch the power method converge.
A_u <- as_adjacency_matrix(G_undir, sparse = FALSE)
N   <- vcount(G_undir)

n <- 20

x <- rep(1, N)
rows <- matrix(NA_real_, nrow = n, ncol = N)
for (i in seq_len(n)) {
  x <- as.numeric(A_u %*% x)
  x <- x / sqrt(sum(x^2))
  rows[i, ] <- round(x, 3)
}

iter_df <- as.data.frame(rows)
colnames(iter_df) <- V(G_undir)$name
rownames(iter_df) <- paste("iter", seq_len(n))

reference <- eigen_centrality(G_undir)$vector
reference <- reference / sqrt(sum(reference^2))
iter_df["igraph", ] <- round(reference, 3)
iter_df
```

### 6.2 PageRank with the power method

For PageRank we

1. row-normalise the adjacency matrix: `A_hat = D^(-1) A`,
2. mix in a teleportation term with damping factor `d = 0.85`: `M = d * A_hat + (1 - d) / N`,
3. iterate `x <- x %*% M` until convergence (note: row-stochastic, so we left-multiply).

```{r}
#| label: c-pagerank
# PageRank is a random walk plus teleportation.
# Small change to try: set `damping <- 0.50` and compare the result.
A <- as_adjacency_matrix(G_dir, sparse = FALSE)
N <- vcount(G_dir)
damping <- 0.85

deg_out <- rowSums(A)
D <- diag(1 / ifelse(deg_out > 0, deg_out, 1))
A_h <- D %*% A

x <- rep(1 / N, N)
for (i in 1:100) {
  x <- as.numeric(x %*% (damping * A_h + (1 - damping) / N))
}

ref <- page_rank(G_dir, damping = damping)$vector
tibble(person = V(G_dir)$name,
       manual = round(x, 4),
       igraph = round(ref, 4))
```

### 6.3 Random walker on an undirected graph ≈ degree centrality

If you let a random walker bounce around an undirected graph forever, the fraction of time it spends at node `i` is proportional to its **degree**. That's why on undirected graphs PageRank (without damping) reduces to degree centrality.

```{r}
#| label: c-randomwalk
# On an undirected graph, long-run random-walk visits track degree.
A <- as_adjacency_matrix(G_undir, sparse = FALSE)
N <- vcount(G_undir)
D <- diag(1 / pmax(rowSums(A), 1))
A_h <- D %*% A

x <- rep(1 / N, N)
for (i in 1:100) {
  x <- as.numeric(x %*% A_h)
}

x <- x / sqrt(sum(x^2))
deg <- as.numeric(degree(G_undir))
deg <- deg / sqrt(sum(deg^2))
tibble(person = V(G_undir)$name,
       `random-walker`       = round(x, 4),
       `degree (normalised)` = round(deg, 4))
```

---

## Section 7: Centrality side-by-side

Different centralities crown different nodes as "most important". To make that concrete, here is the **same** Florentine network drawn four times, with nodes sized by four different centralities. The same Fruchterman-Reingold layout is used in each panel so you can compare positions at a glance.

```{r}
#| label: c-sidebyside
#| fig-width: 10
#| fig-height: 9
set.seed(1)
layout_flo <- layout_with_fr(G_flo)
names_flo <- V(G_flo)$name

centralities <- list(
  degree      = as.numeric(degree(G_flo)),
  betweenness = as.numeric(betweenness(G_flo)),
  closeness   = as.numeric(closeness(G_flo)),
  eigenvector = as.numeric(eigen_centrality(G_flo)$vector)
)

op <- par(mfrow = c(2, 2), mar = c(1, 1, 3, 1))
for (cname in names(centralities)) {
  vals <- centralities[[cname]]
  span <- max(vals) - min(vals)
  if (span < 1e-12) span <- 1
  sizes <- 10 + 22 * (vals - min(vals)) / span

  # Continuous color encoding (coolwarm), consistent with plot_network() above
  colors <- value_to_color(vals)

  # Mark the top node with a thicker frame so it stands out at a glance
  top <- which.max(vals)
  frame_widths <- rep(0.5, vcount(G_flo)); frame_widths[top] <- 3
  frame_colors <- rep("#444444", vcount(G_flo)); frame_colors[top] <- "#191716"

  plot(G_flo, layout = layout_flo,
       vertex.size = sizes, vertex.color = colors,
       vertex.frame.color = frame_colors, vertex.frame.width = frame_widths,
       edge.color = "lightgray",
       vertex.label = names_flo, vertex.label.cex = 0.6,
       vertex.label.color = "#191716",
       main = sprintf("%s  (top: %s)", cname, names_flo[top]))
}
par(op)
mtext("Florentine families - four centralities, one layout", side = 3, line = -1.5,
      outer = TRUE, cex = 0.9)
```

**Take-away.** The Medici rank highest on most centrality measures, echoing their historical dominance. But the Strozzi move up or down depending on the metric: they have many connections and strong local clustering, yet sit off-center in the network, which lowers their betweenness.

Same network, different measures, different "winners" — which is why we use more than one measure of centrality.
